Topological Graph Theory

### 2018-10-02 Topology/Paper/paper.tex
hi

\section{Introduction}
Topological graph theory is an entire field within topology and as such this paper is by no means meant to cover all of topological graph theory in any depth. This paper instead will first cover a rather shallow overview of the field, followed by a more in depth study of graphs and their genus. The overview will mainly be focused on giving a thorough understanding of what topological graph theory is as well as to briefly cover the history of the field. In giving an overview of the field we will cover some of the basic concepts and definitions needed for the more rigorous part of the paper; as this may be confusing, all definitions will be provided in a glossary at the end. After the overview we will dive into Kuratowski's Theorem, We will go through and attempt to have an intuitive understanding of a Kuratowski's Theorem and it's proof. After proving Kuratowski's Theorem, we will continue onto talking about generalizations of the theorem and map colorings, however their coverage will be rather shallow and lacking proofs.

@@

\section{Planar Graphs}

@@

  \node[main node] (1) {$A$};
  \node[main node] (2) [below of=1] {$B$};
  \node[main node] (3) [below right of=2] {$C$};
  \node[main node] (4) [below right of=1] {$D$};

  \path[every node/.style={font=\sffamily\small}]
    (1) edge node {} (4)
    	edge [bend right = 100]node {} (3)
    (2) edge node {} (1)
        edge node {} (4)
    (3) edge node {} (2)
    (4) edge node {} (3);
\end{tikzpicture}
\end{center}

So if this \gls{graph} can be drawn without intersection, can any \gls{graph} be drawn without intersections? If some can and some can't how do we tell which can be drawn and which can not?

\section{Kuratowski's Theorem}
Kuratowski's theorem tell us exactly what \glspl{graph} can be drawn on a plane or sphere. We say either a plane or sphere as it turns out to be the same question. %TODO prove that drawing on plane and sphere is same question
Now before we tackle the theorem let's lay down some terminology.

\subsection{Terminology}

\subsubsection{Terminology: face}

@@

	  \node[main node] (1) {$A$};
	  \node[main node] (2) [below left of=1] {$B$};
	  \node[main node] (3) [below right of=1] {$C$};

	  \path[every node/.style={font=\sffamily\small}]
	  (1) edge node {} (2)
		  edge node {} (3)
	  (2) edge node {} (3);

	\end{tikzpicture}
\end{center}

We say this \gls{graph} has two \glspl{face}. One \gls{face} is easy to see and contained within triangle $ABC$. The other \gls{face} is a bit harder to see, however it is all the space outside of the triangle $ABC$. We may call this second face the outer \gls{face} for this drawing of the \gls{graph}, however it is also possible to draw it such that the inner face may be the outer face. You can think of this as kind of turning the triangle inside out.

A face only makes sense to talk about for a \gls{graph} that can be embedded in a certain space. For example the \gls{graph} $K_4$ may look like it has $5$ faces in the following drawing.

	\path[every node/.style={font=\sffamily\small}]
	(1) edge node {} (4)
	edge node {} (3)
	(2) edge node {} (1)
	edge node {} (4)
	(3) edge node {} (2)
	(4) edge node {} (3);
	\end{tikzpicture}
\end{center}
This would be wrong as a face assumes a drawing without \glspl{edge} intersecting. This can be rectified if the \gls{graph} can be drawn without intersection and thus we find this drawing.
\begin{center}
	\begin{tikzpicture}[auto, node distance=3cm, every loop/.style={},
	thick,main node/.style={circle,draw,font=\sffamily\Large\bfseries}]

	\node[main node] (1) {$A$};
	\node[main node] (2) [below of=1] {$B$};
	\node[main node] (3) [below right of=2] {$C$};
	\node[main node] (4) [below right of=1] {$D$};

	\path[every node/.style={font=\sffamily\small}]
	(1) edge node {} (4)
	edge [bend right = 100]node {} (3)
	(2) edge node {} (1)
	edge node {} (4)
	(3) edge node {} (2)
	(4) edge node {} (3);
	\end{tikzpicture}
\end{center}
This means there are actually $4$ \glspl{face} in $K_4$.

\subsubsection{Terminology: contraction}

Let us consider the following \gls{graph}.

@@

This can then be reduced into a \gls{graph} again by simply treating $b_r$ and $b_s$ as the same edge. This operation is particularly important as we will prove that if a \gls{graph} is planar, then so is any \gls{graph} or \gls{multigraph} obtained by contracting an \gls{edge}. The proof for this will be provided.
%TODO put refrence to proof for contraction of planar graph is planar

\subsubsection{Terminology: subgraph}
We can say that a \gls{graph} $A$ is a \gls{subgraph} of another \gls{graph} $B$, if all \glspl{vertex} in $A$ are also in $B$ and each \glspl{edge} in $A$ is also in $B$.

\subsection{Groundwork}
Here we would like to prove some lemmas that we would like to use while proving Kuratowski's theorem.

Now for some extensions of Kuratowski's theorem

@@

This is not needed for our proof of Kuratowski's theorem, however it does help to think about drawing graphs on planes sometimes as well as helping to think about extensions beyond Kuratowski's theorem.

>>>>

\section{Introduction}
Topological graph theory is an entire field within topology and as such this paper is by no means meant to cover all of topological graph theory in any depth. This paper instead will first cover a rather shallow overview of the field, followed by a more in depth study of graphs and their genus. The overview will mainly be focused on giving a thorough understanding of what topological graph theory is as well as to briefly cover the history of the field. In giving an overview of the field we will cover some of the basic concepts and definitions needed for the more rigorous part of the paper. After the overview we will dive into Kuratowski's Theorem, We will go through and attempt to have an intuitive understanding of a Kuratowski's Theorem and it's proof. After proving Kuratowski's Theorem, we will continue onto talking about generalizations of the theorem and map colorings, however their coverage will be rather shallow and lacking proofs.

@@

\section{Graph Theory Background}

\subsection{Planer Graphs}

@@

  \node[main node] (1) {$A$};
  \node[main node] (2) [above left of=1] {$B$};
  \node[main node] (3) [below left of=1] {$C$};
  \node[main node] (4) [right of=1] {$D$};

So if this \gls{graph} can be drawn without intersection, can any \gls{graph} be drawn without intersections? If some can and some can't how do we tell which can be drawn and which can not? The answer to this comes in Kuratowski's theorem, however before we can even state this theorem we need to build up a bit of terminology for graph theory.

\subsection{Graph Drawing}
A graph drawing is exactly what it sounds like. We've already seen drawings of graphs like the one for K\"oingsberg and two drawings of $k_4$, this means there may be multiple distinct drawings for the same graph. Now we don't need to be very worried about what defines a drawing, that won't be important to us. Simply think of it as the drawing.

If a drawing has no edges intersecting it is said to be a plane drawing. If there exists such a drawing for a particular graph, then that graph is a planar graph.

\subsection{Face}
Faces are a bit of an odd property here as they fundamentally are actually properties only of plane drawings and not graphs themselves; however the number of faces, as we will see stays consistent between any plane drawing of a planar graph and as such the number of faces is a property that planar graphs have.

A face is defined as a connected space that contains no edges or verticies and itself is bounded by edges and verticies.

If you think back to high-school geometry and cubes you may recall that a each of the corners is a vertex, the lines connecting verticies are edges and the area between the edges are faces. In a graph we have verticies connected by edges and when we draw them there are empty areas enclosed by edges and verticies. For example the following would correspond with a cube
\begin{center}
	\begin{tikzpicture}[auto, node distance=3cm, every loop/.style={},
	thick,main node/.style={circle,draw,font=\sffamily\Large\bfseries}]

	\node[main node] (1) {$A$};
	\node[main node] (2) [below of=1] {$B$};
	\node[main node] (3) [right of=1] {$C$};
	\node[main node] (4) [right of=2] {$D$};
	\node[main node] (5) [above left of=1] {$E$};
	\node[main node] (6) [below left of=2] {$F$};
	\node[main node] (7) [above right of=3] {$G$};
	\node[main node] (8) [below right of=4] {$H$};

	\path[every node/.style={font=\sffamily\small}]
	(1) edge node {} (2)
	edge node {} (3)
	edge node {} (5)
	(2) edge node {} (4)
	edge node {} (6)
	(3) edge node {} (4)
	edge node {} (7)
	(4) edge node {} (8)
	(5) edge node {} (6)
	edge node {} (7)
	(6) edge node {} (8)
	(7) edge node {} (8);
	\end{tikzpicture}
\end{center}
Now we can imagine that if we look at the cube straight on maybe $ABDC$ would be the face we are looking at. So too we see the area enclosed in $ABDC$ is a face by the definition we gave. The same is true for $ABFE$, $ACGE$, $DBFH$, and $DCGH$, however that leaves us with only five faces on this cube. The last face must logically come from $EFHG$, however the area enclosed by that contains all the other edges and verticies, so it doesn't fit our definition. However we may notice that the infinite space outside $EFHG$ does not contain any verticies, and therefore we get a sixth face.

Consider then the following very simple graph:
\begin{center}
	\begin{tikzpicture}[auto, node distance=3cm, every loop/.style={},
	thick,main node/.style={circle,draw,font=\sffamily\Large\bfseries}]

	\node[main node] (1) {$A$};
	\node[main node] (2) [right of=1] {$B$};

	\end{tikzpicture}
\end{center}
Now we still only have one face, however it doesn't have as nice a boundary as they did in the cube. However if we look at all of space excluding $A$ and $B$ then we still get a valid space by our definition.

\subsection{Path}
Now moving onto a more traditional graph property we have the concept of a path. A path is defined as a finite sequence of verticies with the property that each element of the sequence (excluding the last one) has an edge from it to the next element. For example consider the following graph.

@@

	\node[main node] (a) {$A$};
	\node[main node] (b) [below left of=a] {$B$};
	\node[main node] (c) [below right of=a] {$C$};
	\node[main node] (d) [below of=b] {$D$};
	\node[main node] (e) [below of=c] {$E$};
	\node[main node] (f) [above of=a] {$F$};
	\node[main node] (g) [above left of=b] {$G$};
	\node[main node] (h) [above right of=c] {$H$};
	\node[main node] (i) [below left of=d] {$I$};
	\node[main node] (j) [below right of=e] {$J$};

	\path[every node/.style={font=\sffamily\small}]
	(a) edge node {} (b)
	edge node {} (c)
	edge node {} (f)
	(b) edge node {} (d)
	edge node {} (g)
	(c) edge node {} (e)
	edge node {} (h)
	(d) edge node {} (i)
	(e) edge node {} (j)
	(f) edge node {} (g)
	edge node {} (h)
	(g) edge node {} (i)
	(h) edge node {} (j)
	(i) edge node {} (j);
	\end{tikzpicture}
\end{center}

Now the sequence $BGF$ is a path as $B$ connects to $G$ and $G$ connects to $F$. The sequence $ABGFACEJHCABA$ is also a path, however $JEDI$ is not a path as $E$ has no edge to $D$. Notice that how we draw the graph has nothing to do with what is and is not a path.

\subsection{Connected}
A graph is said to be connected if for any pair of verticies, $\pair ab$ there is a path from $a$ to $b$. So if we consider the following graphs 
\begin{center}
	\begin{tikzpicture}[auto, node distance=3cm, every loop/.style={},
	thick,main node/.style={circle,draw,font=\sffamily\Large\bfseries}]

	\node[main node] (1) {};
	\node[main node] (2) [right of=1] {};
	\node[main node] (3) [below of=1] {};
	\node[main node] (4) [right of=3] {};

	\path[every node/.style={font=\sffamily\small}]
	(1)	edge node {} (3)
	(4) edge node {} (1);
	\end{tikzpicture}\quad\quad\quad
	\begin{tikzpicture}[auto, node distance=3cm, every loop/.style={},
	thick,main node/.style={circle,draw,font=\sffamily\Large\bfseries}]

	\node[main node] (1) {};
	\node[main node] (2) [right of=1] {};
	\node[main node] (3) [below of=1] {};
	\node[main node] (4) [right of=3] {};

	\path[every node/.style={font=\sffamily\small}]
	(1)	edge node {} (3)
	(2) edge node {} (3)
	(4) edge node {} (1);
	\end{tikzpicture}\quad\quad\quad
	\begin{tikzpicture}[auto, node distance=3cm, every loop/.style={},
	thick,main node/.style={circle,draw,font=\sffamily\Large\bfseries}]

	\node[main node] (1) {};
	\node[main node] (2) [right of=1] {};
	\node[main node] (3) [right of=2] {};
	\node[main node] (4) [below of=1] {};
	\node[main node] (5) [right of=4] {};
	\node[main node] (6) [right of=5] {};

	\path[every node/.style={font=\sffamily\small}]
	(1)	edge node {} (5)
	(2) edge node {} (4)
	edge node {} (6)
	(3) edge node {} (5);
	\end{tikzpicture}
\end{center}
We find that only the graph in the middle is connected. For the graph on the right consider any vertex on the bottom, there is no path to the vertex above it. The graph on the left is has the upper right vertex isolated from the rest of the graph.

Any graph, connected or not, can be broken into connected components. To do this we simply take a vertex and every other vertex connected to it and call that one component, and then repeat with a vertex not in that component. This sort of breaking apart is nice as often we will prove things about connected graphs that are true about all graphs. For example if all components of a graph are planar then the entire graph must be planar, this will be proven below and it allows us to only deal with connected graphs.

\subsection{Contraction}
This is not a property, but rather an operation or action that we preform on a graph. Let us consider the following \gls{graph}.

@@

This can then be reduced into a \gls{graph} again by simply treating $b_r$ and $b_s$ as the same edge. This operation is particularly important as we will prove that if a \gls{graph} is planar, then so is any \gls{graph} or \gls{multigraph} obtained by contracting an \gls{edge}.

\newpage

\section{Kuratowski's Theorem}
\begin{theorem}[Kuratowski's Theorem]
	A graph $G$ is planar if and only if it does not contain any subgraph $H$ such that a some series of edge contractions on $H$ would produce $K_5$ or $K_{3,3}$.
\end{theorem}

Here $K_5$ refers to the complete graph on 5 verticies, and $K_{3,3}$ is the complete bipartite graph with three verticies in both partitions. Drawings of both theses graphs are below with ($K_5$ on left and $K_{3,3}$ on right).
\begin{center}
	\begin{tikzpicture}[auto, node distance=3cm, every loop/.style={},
	thick,main node/.style={circle,draw,font=\sffamily\Large\bfseries}]

	\node[main node] (1) {};
	\node[main node] (2) [below left of=1] {};
	\node[main node] (3) [below right of=1] {};
	\node[main node] (4) [below of=2] {};
	\node[main node] (5) [below of=3] {};

	\path[every node/.style={font=\sffamily\small}]
	(1)	edge node {} (2)
	edge node {} (3)
	edge node {} (4)
	edge node {} (5)
	(2) edge node {} (3)
	edge node {} (4)
	edge node {} (5)
	(3) edge node {} (4)
	edge node {} (5)
	(4) edge node {} (5);
	\end{tikzpicture}
	\begin{tikzpicture}[auto, node distance=3cm, every loop/.style={},
	thick,main node/.style={circle,draw,font=\sffamily\Large\bfseries}]

	\node[main node] (1) {};
	\node[main node] (2) [below of=1] {};
	\node[main node] (3) [below of=2] {};
	\node[main node] (4) [right of=1] {};
	\node[main node] (5) [below of=4] {};
	\node[main node] (6) [below of=5] {};

	\path[every node/.style={font=\sffamily\small}]
	(1)	edge node {} (4)
	edge node {} (5)
	edge node {} (6)
	(2)	edge node {} (4)
	edge node {} (5)
	edge node {} (6)
	(3)	edge node {} (4)
	edge node {} (5)
	edge node {} (6);
	\end{tikzpicture}
\end{center}
\subsection{Groundwork}

Here we prove some lemmas that we will use while proving Kuratowski's theorem before getting into the meat of the theorem.

@@

This means that we may freely interchange usage of a sphere with a plane; for our purposes they are the same.

\begin{theorem}[Euler's equation for planar graphs]
	Let the number of edges in the graph be $e$, the number of verticies by $v$ and the number of faces be $f$.
\end{theorem}
